### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/revision/response.tex

{\parindent=20pt \it

\item{(9)} Your binary search results in a particular set of
  derivatives, for which the local quintic piece is guaranteed to be
  monotone, but it is clearly not a unique choice. Alternatively, one
  could try to find, among all possible sets of derivatives and
  corresponding monotone quintic interpolants, the best, according to
  some criterion, e.g. least L2 norm. This would turn the problem into
  a constrained optimization problem, which is probably harder to
  solve, but it might, at the same time, give better results and
  overcome the limitation shown in Fig. 1. More motivating details
  would be appreciated.

}

This was a point of our own severe consideration before submitting the
manuscript and this comment reflected our original thoughts. In due
response to this we've implemented naive $L2$ minimization algorithm
over the second derivative that uses gradient descent over the spline
representation to enforce local monotonicity constraints in a
Lagrangian fashion (with an adaptive multiplier on monotonicity
violations). The results we achieved were quite mixed and we hestiate
to incorporate commentary from this excursion into the manuscript. In
truth, formally expanding on this trajectory would require an entirely
new manuscript with deeper study of the problem, while it is not
immediately obvious that the results would be {\it better} than the
present approach. The global $L^2$ minimizer of the second derivative
is simply the unique $C^2$ cubic spline interpolant, but that is known
to produce large and unnecessary local variations in the approximation
(along with not being shape preserving). Included in this response (to
reviewers only) are some of the preliminary visuals we produced in
exploring this approach. This is by no means final and conclusive, but
the results were mixed enough for us to warrant not including them in
the paper.

@@

\item{(12)} The output are values -- why not the spline coefficients? If
  the output are values one would hope for a visualization package.

}

The code includes suitable documentation and examples to conform to
any external visualization package desired by a user. Given the wide
variety of mechanisms by which a user might want to visualize or use
the results of this code, it is nearly impossible to provide a good
generic solution to visualization.

For the convenience of reviewers, we are providing the code with which
we generated the visuals for this manuscript. Although we note it has
very specific system requirements to operate effectively. Namely: (1)
running a POSIX shell (on macOS, Ubuntu, $\ldots$); (2) having {\tt
  python3} of at least version $3.7$ and the packages {\tt numpy},
{\tt tlux}, and {\tt rpy2}; (3) having installed {\tt gfortran} (or
otherwise reconfiguring the codes to use another Fortran compiler of
choice); and (4) having installed {\tt R} (for Schumaker splines). The
Fortran codes associated with this work as well as those from TOMS 574
(quadratic $C^1$) and BVSPIS from TOMS 770 are all wrapped to be
callable directly from {\tt python3}, which will be the parent process
for all calls. A simple test usage can be seen in the provided {\tt
  python3} file {\tt test.py}. To be abundantly clear, this code is
{\bf not} intended for review and publication, but rather convenient
access to various codes for reviewers to use for visualizing and
comparing results.

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Finished review draft, sending to coauthors.

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations whose
second derivative functions often have large $L^2$ norm (approximations
with large {\it wiggle}). In an attempt to capture the local shape of
the data while minimizing wiggle, this package uses a facet model from
image processing [Haralick and Watson 1981] to estimate first and
second derivatives at breakpoints. Rather than picking a local linear
or quadratic fit with minimal residual, this work uses a quadratic
facet model that selects the local quadratic interpolant with minimum
magnitude curvature.

@@

\nobreak
where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
\le n$, and $n \ge 3$. Returns the slope and curvature at $X_i$ of the
local quadratic interpolant with minimum magnitude curvature.

@@

\item{} {\tt else if} $\bigl(1<i<n\wedge (Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0
\bigr)$ {\tt then}
\itemitem{} The point $(X_i$, $Y_i)$ is an extreme point. The
quadratic with minimum magnitude
curvature that has slope zero at $X_i$ will be the facet chosen.
\itemitem{} $f_1\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$, $(X_i,Y_i)$,
  and $Df_1(X_i) = 0$.
\itemitem{} $f_2\:\hbox{interpolant to }(X_i,Y_i)$, $(X_{i+1},Y_{i+1})$,
  and $Df_2(X_i) = 0$.
\itemitem{} {\tt if} $\bigl(|D^2f_1| \leq |D^2f_2|\bigr)$ {\tt then
  return} $\bigl(0$, $D^2f_1\bigr)$
\itemitem{} {\tt else return} $\bigl(0$, $D^2f_2\bigr)$
\itemitem{} {\tt endif}

\item{} {\tt else}
\itemitem{} {The point $(X_i$, $Y_i)$ is in a monotone segment of
data. In the following, it is possible that $f_1$ or $f_3$ do
not exist because $i \in \{2, n-1\}$. In those cases, the minimum
magnitude curvature among existing quadratics is chosen.}
\itemitem{} $f_1\:\hbox{interpolant to }(X_{i-2},Y_{i-2})$,
  $(X_{i-1},Y_{i-1})$, and $(X_i,Y_i)$.
\itemitem{} $f_2\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$,
  $(X_i,Y_i)$, and $(X_{i+1},Y_{i+1})$.
\itemitem{} $f_3\:\hbox{interpolant to }(X_i,Y_i)$,
  $(X_{i+1},Y_{i+1})$, and $(X_{i+2},Y_{i+2})$.

@@

\nobreak
where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$.

@@

\item{} {The necessity of Steps 1--4 follows directly from the
  fact that $f$ is $C^2$. The following Steps 7--13 coincide with a
  simplified condition for quintic monotonicity that reduces to one of
  cubic positivity studied by Schmidt and He{\ss} [1988]. Given
  $\alpha$, $\beta$, $\gamma$, and $\delta$ as defined by Schmidt and
  He{\ss}, monotonicity results when $\alpha \geq 0$, $\delta \geq 0$,
  $\beta \geq \alpha - 2 \sqrt{\alpha \delta}$, and $\gamma \geq
  \delta - 2 \sqrt{\alpha \delta}$.  Step 4 checked for $\delta < 0$,
  Step 8 checks $\alpha < 0$, Step 10 checks $\beta < \alpha - 2
  \sqrt{\alpha \delta}$, and Step 11 checks $\gamma < \delta - 2
  \sqrt{\alpha \delta}$. If none of the monotonicity conditions are
  violated, then the order six piece is monotone and Step 12
  concludes.}

@@

Given the minimum magnitude curvature nature of the initial derivative
estimates, it is desirable to make the smallest necessary changes to
the initial interpolating spline $Q$ while enforcing monotonicity. In
practice a binary search for the boundary of monotonicity is used in
place of solely shrinking $DQ$ and $D^2Q$ at breakpoints adjoining
{\it active\/} intervals: intervals over which $Q$ is nonmonotone at
least once during the search. The binary search considers a Boolean
function $B_i(s)$, for $0 \le s \le 1$, that is true if the order six
polynomial piece of $Q(x)$ on $[X_i, X_{i+1}]$ matching derivatives
$Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false otherwise. 
The binary search is only applied at those breakpoints adjoining intervals
$[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and hence $B_i(1)$ is false.
It is further assumed that there exists $0 \le s^* \le 1$ such that $B_i(s)$
is true for $0 \le s \le s^*$ and false for some $1 > s > s^*$. Since the
derivative conditions at interior breakpoints are shared by intervals left
and right of the breakpoint, the binary search is performed at all
breakpoints simultaneously.  Specifically, the monotonicity of $Q$ is
checked on all active intervals in each step of the binary search to
determine the next derivative modification at each breakpoint. The goal
of this search is to converge on the boundary of the monotone region in
the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
Watson [1994]) for all intervals. This multiple-interval binary search
allows the value zero to be obtained for all (first and second) derivative
values in a fixed maximum number of computations, hence has no effect on
computational complexity order. This binary search algorithm is outlined
below.

>>>>

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations whose
second derivative functions often have large \red{local} $L^2$ norm
(approximations with large {\it wiggle}). In an attempt to capture the
local shape of the data while minimizing wiggle, this package uses a
facet model from image processing [Haralick and Watson 1981] to
estimate first and second derivatives at breakpoints. Rather than
picking a local linear or quadratic fit with minimal residual, this
work uses a quadratic facet model that selects the local quadratic
interpolant with minimum magnitude \red{second derivative}.

@@

\nobreak
where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
\le n$, and $n \ge 3$. Returns the \red{first and second derivative} at 
$X_i$ of the local quadratic interpolant with minimum magnitude
second derivative. \red{Approximate equality is denoted with
$\approx$ and considers two numbers to be equal once they are within
the machine precision $\epsilon$ of each other.}

@@

\item{} {\tt else if} $\bigl(1<i<n\wedge (Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0
\bigr)$ {\tt then}
\itemitem{} The point $(X_i$, $Y_i)$ is an extreme point. The
quadratic with minimum magnitude \red{second derivative} that has slope
zero at $X_i$ will be the facet chosen.
\itemitem{} $f_1\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$, $(X_i,Y_i)$,
  and $Df_1(X_i) = 0$.
\itemitem{} $f_2\:\hbox{interpolant to }(X_i,Y_i)$, $(X_{i+1},Y_{i+1})$,
  and $Df_2(X_i) = 0$.
\itemitem{} {\tt if} $\bigl(|D^2f_1| \leq |D^2f_2|\bigr)$ {\tt then
  return} $\bigl(0$, $D^2f_1\bigr)$
\itemitem{} {\tt else return} $\bigl(0$, $D^2f_2\bigr)$
\itemitem{} {\tt endif}

\item{} {\tt else}
\itemitem{} {The point $(X_i$, $Y_i)$ is in a monotone segment of
data. In the following, it is possible that $f_1$ or $f_3$ do
not exist because $i \in \{2, n-1\}$. In those cases, the minimum
magnitude \red{second derivative} among existing quadratics is chosen.}
\itemitem{} $f_1\:\hbox{interpolant to }(X_{i-2},Y_{i-2})$,
  $(X_{i-1},Y_{i-1})$, and $(X_i,Y_i)$.
\itemitem{} $f_2\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$,
  $(X_i,Y_i)$, and $(X_{i+1},Y_{i+1})$.
\itemitem{} $f_3\:\hbox{interpolant to }(X_i,Y_i)$,
  $(X_{i+1},Y_{i+1})$, and $(X_{i+2},Y_{i+2})$.

@@

\nobreak

where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$. \red{Approximate equality is denoted with
$\approx$ and considers two numbers to be equal once they are within
the machine precision $\epsilon$ of each other.}

@@

\item{} {The necessity of Steps 1--4 follows directly from the
  fact that $f$ is $C^2$. The following Steps 7--13 coincide with a
  simplified condition for quintic monotonicity that reduces to one of
  cubic positivity studied by Schmidt and He{\ss} [1988]. Given
  $\alpha$, $\beta$, $\gamma$, and $\delta$ as defined by Schmidt and
  He{\ss}, monotonicity results when $\alpha \geq 0$, $\delta \geq 0$,
  $\beta \geq \alpha - 2 \sqrt{\alpha \delta}$, and $\gamma \geq
  \delta - 2 \sqrt{\alpha \delta}$.  Step 4 checked for $\delta < 0$,
  Step 8 checks $\alpha < 0$, Step 10 checks $\beta < \alpha - 2
  \sqrt{\alpha \delta}$, and Step 11 checks $\gamma < \delta - 2
  \sqrt{\alpha \delta}$. If none of the monotonicity conditions are
  violated, then the \red{degree five} piece is monotone and Step 12
  concludes.}

@@

Given the minimum \red{curvature (minimum magnitude of the second derivative)}
nature of the initial derivative estimates, it is desirable to make
the smallest necessary changes to the initial interpolating spline $Q$
while enforcing monotonicity. In practice a binary search for the
boundary of monotonicity is used in place of solely shrinking $DQ$ and
$D^2Q$ at breakpoints adjoining {\it active\/} intervals: intervals
over which $Q$ is nonmonotone at least once during the search. The
binary search considers a Boolean function $B_i(s)$, for $0 \le s \le
1$, that is true if the order six polynomial piece of $Q(x)$ on
$[X_i, X_{i+1}]$ matching derivatives
$Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false
otherwise.  The binary search is only applied at those breakpoints
adjoining intervals $[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and
hence $B_i(1)$ is false.  It is further assumed that there exists
$0 \le s^* \le 1$ such that $B_i(s)$ is true for $0 \le s \le s^*$ and
false for some $1 > s > s^*$. Since the derivative conditions at
interior breakpoints are shared by intervals left and right of the
breakpoint, the binary search is performed at all breakpoints
simultaneously.  Specifically, the monotonicity of $Q$ is checked on
all active intervals in each step of the binary search to determine
the next derivative modification at each breakpoint. The goal of this
search is to converge on the boundary of the monotone region in the
$(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
Watson [1994]) for all intervals. This multiple-interval binary search
allows the value zero to be obtained for all (first and second)
derivative values in a fixed maximum number of computations, hence has
no effect on computational complexity order. This binary search
algorithm is outlined below.

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Finished review draft, sending to coauthors.

% Color definitions.
\input colordvi
\def\BEGINC{\textPeach} \def\ENDC{\textBlack}
\def\red#1{\BEGINC#1\ENDC}
\let\beginred=\BEGINC
\let\endred=\ENDC

@@

% Abstract and Introduction
{\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
monotone quintic spline interpolants to univariate monotone data. Using sharp
theoretical monotonicity constraints, first and second derivative estimates
at data provided by a quadratic facet model are refined to produce a
univariate $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
details, complexity and sensitivity analyses, usage information, a
brief performance study, and comparisons with other spline approaches are included.

>>>>

% Color definitions.
\input colordvi
\def\BEGINC{\textRed}
\def\ENDC{\textBlack}
\def\red#1{\BEGINC#1\ENDC}
\let\beginred=\BEGINC
\let\endred=\ENDC

@@

% Abstract and Introduction
{\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
monotone quintic spline interpolants to \red{univariate} monotone data. Using sharp
theoretical monotonicity constraints, first and second derivative estimates
at data provided by a quadratic facet model are refined to produce a
\red{univariate} $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
details, complexity and sensitivity analyses, usage information, a
brief performance study, and comparisons with other spline approaches are included.
